---
title: "Examples from US EPA's Unified Guidance"
author: "Stephen P. Millard"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Reproduce Examples in Chapter 6 of the EPA Guidance Document

Unless otherwise noted, differences between what is shown in the EPA Guidance Document and results shown here are due to the EPA Guidance Document rounding intermediate results prior to the final result.

```{r}
library(EnvStats)
```
# Figure-6-2, p. 6-15
```{r figure-6-2}
plotPredIntNormTestPowerCurve(n = 10, k = 1, 
	range.delta.over.sigma = c(0, 5), 
	pi.type = "upper", conf.level = 0.99, 
	ylim = c(0, 1), 
	xlab = expression(paste(Delta, " (SDs above Background)")), 
	ylab = "Power", cex.main = 1.1, 
	main = "Figure 6-2. Normal Power Curve (n = 10) for 99% Prediction Limit Test")
```

# Figure 6-3, p. 6-17
```{r figure-6-3}
plotPredIntNormTestPowerCurve(n = 10, k = 1, 
	range.delta.over.sigma = c(0, 5), 
	pi.type = "upper", conf.level = 0.99, 
	ylim = c(0, 1), 
	xlab = "SD Units Above BG", 
	ylab = "Power",
	main = "Figure 6-3. EPA Reference Power Curves")

plotPredIntNormTestPowerCurve(n = 10, k = 2, conf.level = 0.99, 
	add = TRUE, plot.col = "red", plot.lty = 2)

plotPredIntNormTestPowerCurve(n = 10, k = 4, conf.level = 0.99, 
	add = TRUE, plot.col = "blue", plot.lty = 3)

legend("topleft", c("Quarterly", "Semi-Annual", "Annual"), lty = 3:1, 
	lwd = 3 * par("cex"), col = c("blue", "red", "black")) 
```
# Example 6-1, p. 6-18
```{r figure-6-1}
predIntNormTestPower(n = 10, k = 1, 
	delta.over.sigma = (15 - 6) / 2, 
	pi.type = "upper", conf.level = 0.99)
```

# Example 6-3, pp. 6-20 to 6-21
```{r example-6-3}
stats   <- summaryStats(Sulfate.ppm ~ 1, data = EPA.09.Ex.6.3.sulfate.df,
	data.name = "Sulfate", digits = 1)
stats

n     <- stats[, "N"]
SD    <- stats[, "SD"]
delta <- c(0, 200)

plotPredIntNormTestPowerCurve(n = n, k = 1, 
	axes = F,
	range.delta.over.sigma = range(delta / SD), 
	pi.type = "upper", conf.level = 0.99, 
	ylim = c(0, 1), 
	xlab = "Sulfate Conc. Increase (ppm)", 
	ylab = "Power",
	main = "Figure 6-3. Approximate s-Based Power Curve for Sulfate")
axis(2)
x.axis.ticks <- seq(0, 200, by = 50)
axis(1, at = x.axis.ticks / SD, labels = x.axis.ticks)
box()
abline(v = 75 / SD, lty = 2, lwd = 2)

predIntNormTestPower(n = n, k = 1, delta.over.sigma = 75 / SD, 
	pi.type = "upper", conf.level = 0.99)

rm(stats, n, SD, delta, x.axis.ticks)
```


# Reproduce Examples in Chapter 7 of the EPA Guidance Document

# Example 7-1, pp. 7-26 to 7-28
```{r example-7-1}
stats <- summaryStats(Arsenic.ug.per.l ~ Data.Source, 
	data = EPA.09.Ex.7.1.arsenic.df,
	combine.groups = F, digits = 2, stats.in.rows = TRUE)
stats

# 1-of-2 future values option
kappa <- predIntNormSimultaneousK(n = 8, k = 1, m = 2, r = 6, 
	rule = "k.of.m", pi.type = "upper", 
	conf.level = 0.95)
kappa <- round(kappa, 2)
kappa

# NOTE:  EPA Guidance Document states kappa = 1.83


Arsenic <- EPA.09.Ex.7.1.arsenic.df$Arsenic.ug.per.l
Source <- EPA.09.Ex.7.1.arsenic.df$Data.Source

UPL <- predIntNormSimultaneous(x = Arsenic[Source == "Historical"], 
	k = 1, m = 2, r = 6, 
	rule="k.of.m", pi.type="upper", conf.level = 0.95)$interval$limits["UPL"]
UPL <- round(UPL, 2)
UPL

# NOTE:  EPA Guidance Document states UPL = 70.2 ug/l


#####
# Upper Tolerance Limit
UTL <- tolIntNorm(x = Arsenic[Source == "Historical"], 
	coverage = 0.95, cov.type = "content", 
	ti.type = "upper", conf.level = 0.95)$interval$limits["UTL"]

UTL <- round(UTL, 1)
UTL

#####
# 90% and 95% Confidence Limits for the Mean for Case 1 and Case 2

# Case 1
ci.lower.limit.90.Pct <- enorm(Arsenic[Source == "Case.1"], ci = T, 
	ci.type = "lower", conf.level = 0.9)$interval$limits["LCL"]
ci.lower.limit.95.Pct <- enorm(Arsenic[Source == "Case.1"], ci = T, 
	ci.type = "lower", conf.level = 0.95)$interval$limits["LCL"]
ci.upper.limit.90.Pct <- enorm(Arsenic[Source == "Case.1"], ci = T, 
	ci.type = "upper", conf.level = 0.9)$interval$limits["UCL"]
ci.upper.limit.95.Pct <- enorm(Arsenic[Source == "Case.1"], ci = T, 
	ci.type = "upper", conf.level = 0.95)$interval$limits["UCL"]
Case.1.limits <- c(ci.lower.limit.90.Pct, ci.lower.limit.95.Pct, 
	ci.upper.limit.90.Pct, ci.upper.limit.95.Pct)

# Case 2
ci.lower.limit.90.Pct <- enorm(Arsenic[Source == "Case.2"], ci = T, 
	ci.type = "lower", conf.level = 0.9)$interval$limits["LCL"]
ci.lower.limit.95.Pct <- enorm(Arsenic[Source == "Case.2"], ci = T, 
	ci.type = "lower", conf.level = 0.95)$interval$limits["LCL"]
ci.upper.limit.90.Pct <- enorm(Arsenic[Source == "Case.2"], ci = T, 
	ci.type = "upper", conf.level = 0.9)$interval$limits["UCL"]
ci.upper.limit.95.Pct <- enorm(Arsenic[Source == "Case.2"], ci = T, 
	ci.type = "upper", conf.level = 0.95)$interval$limits["UCL"]
Case.2.limits <- c(ci.lower.limit.90.Pct, ci.lower.limit.95.Pct, 
	ci.upper.limit.90.Pct, ci.upper.limit.95.Pct)

mat <- round(rbind(Case.1.limits, Case.2.limits), 1)
dimnames(mat) <- list(c("Case.1", "Case.2"), 
	c("90%.LCL", "95%.LCL", "90%.UCL", "95%.UCL"))
mat

rm(Arsenic, Source, stats, kappa, UPL, UTL, 
	ci.lower.limit.90.Pct, ci.lower.limit.95.Pct,
	ci.upper.limit.90.Pct, ci.upper.limit.95.Pct,
	Case.1.limits, Case.2.limits, mat)
```

# Reproduce Examples in Chapter 9 of the EPA Guidance Document

# Example 9-1, pp. 9-3 to 9-5
```{r}
stats <- summaryStats(TCE.mg.per.L ~ Well, 
	data = EPA.09.Table.9.1.TCE.df,
	combine.groups = F, 
	digits = 3, stats.in.rows = TRUE)
stats

Min.Max.mat <- stats[c("Min", "Max"), ]
Min.Max.mat

TCE            <- EPA.09.Table.9.1.TCE.df$TCE.mg.per.L
Date           <- EPA.09.Table.9.1.TCE.df$Date
Well           <- EPA.09.Table.9.1.TCE.df$Well
Data.Qualifier <- EPA.09.Table.9.1.TCE.df$Data.Qualifier

index.Well.1.Min <- Well == "Well.1" & !is.na(TCE) & TCE == Min.Max.mat["Min", "Well.1"]
index.Well.1.Max <- Well == "Well.1" & !is.na(TCE) & TCE == Min.Max.mat["Max", "Well.1"]
index.Well.2.Min <- Well == "Well.2" & !is.na(TCE) & TCE == Min.Max.mat["Min", "Well.2"]
index.Well.2.Max <- Well == "Well.2" & !is.na(TCE) & TCE == Min.Max.mat["Max", "Well.2"]

DQ.vec <- c(paste(Data.Qualifier[index.Well.1.Min], collapse = ","), 
	paste(Data.Qualifier[index.Well.1.Max], collapse = ","),
	paste(Data.Qualifier[index.Well.2.Min], collapse = ","), 
	paste(Data.Qualifier[index.Well.2.Max], collapse = ","))
DQ.mat <- matrix(DQ.vec, 2, 2)
DQ.mat

df <- data.frame(Min.Max.mat[, 1], DQ.mat[, 1], Min.Max.mat[, 2], DQ.mat[, 2])
row.names(df) <- c("Min", "Max")
names(df) <- c("Well.1.TCE", "Well.1.DQ", "Well.2.TCE", "Well.2.DQ")
df

Well.index <- Well == "Well.1"
ND.index   <- Data.Qualifier == "U"
NA.index   <- is.na(TCE)

plot(Date[Well.index], TCE[Well.index], type = "n", 
	ylim = c(0, max(TCE, na.rm = T)),
	xlab = "Time", ylab = "Trichloroethene (mg/L)")
points(Date[Well.index & !ND.index], TCE[Well.index & !ND.index], 
	cex = 2, pch = 16, col = "blue")
points(Date[Well.index & ND.index], TCE[Well.index & ND.index], 
	cex = 2, pch = 1, col = "blue")
lines(Date[Well.index & !NA.index], TCE[Well.index & !NA.index], 
	lwd = 2, col = "blue")

Well.index <- Well == "Well.2"
points(Date[Well.index & !ND.index], TCE[Well.index & !ND.index], 
	cex = 2, pch = 15, col = "red")
points(Date[Well.index & ND.index], TCE[Well.index & ND.index], 
	cex = 2, pch = 14, col = "red")
lines(Date[Well.index & !NA.index], TCE[Well.index & !NA.index], 
	lwd = 2, lty = 2, col = "red")

legend(Date[1], 0.25, legend = c("Well 1", "Well 2"),
	col = c("blue", "red"), lty = c(1, 2), lwd = rep(2, 2),
	pch = c(16, 15), horiz = T)

title(main = "Figure 9-1. Time Series Plot of Trichloroethene Groudwater\nfor Wells 1 and 2 from 2005-2007")

rm(TCE, Date, Well, Data.Qualifier, stats, Min.Max.mat, 
	index.Well.1.Min, index.Well.1.Max, 
	index.Well.2.Min, index.Well.2.Max, 
	DQ.vec, DQ.mat, df, Well.index, ND.index, NA.index) 
```

# Example 9-2, pp. 9-7 to 9-8, Using Standard Boxplot

 NOTE: Instructions for producing Figure 9-2 include
       "Step 6. Draw the whiskers from each end of the 
       box to the furthest data point to show the 
       full range of the data."

       This is not how boxplots are conventionally drawn,
       so instead the standard boxplot is drawn here, where
       the lower whisker extends to the smallest value within
       1.5 x IQR of the 25th Percentile and the upper whisker
       extends to to the largest value within 1.5 x IQR of 
       the 75th Percentile.

       To produce a boxplot with whiskers that exetend to
       the extreme values, set the argument range=0 in the
       call to the boxplot() function.
```{r}
stats <- summaryStats(TCE.mg.per.L ~ Well, 
	data = EPA.09.Table.9.1.TCE.df,
	combine.groups = F, digits = 3, stats.in.rows = TRUE)
stats

stats.no.ND <- summaryStats(TCE.mg.per.L ~ Well, 
	data = EPA.09.Table.9.1.TCE.df, 
	subset = Data.Qualifier != "U",
	combine.groups = F, digits = 3, stats.in.rows = TRUE)

stats.no.ND

box.list <- boxplot(TCE.mg.per.L ~ Well, data = EPA.09.Table.9.1.TCE.df,
	boxwex = 0.3, varwidth = T,
	ylab = "Trichloroethene (mg/L)",
	main = "Figure 9-2. Boxplots of Trichloroethene Data for Wells 1 & 2")

points(1:2, stats["Mean",], pch = 8, cex = 0.7)

TCE            <- EPA.09.Table.9.1.TCE.df$TCE.mg.per.L
Well           <- EPA.09.Table.9.1.TCE.df$Well
limits <- sapply(split(TCE, Well), 
	function(x) enorm(x, ci=T)$interval$limits)

points(1:2, limits["UCL", ], pch = 94, cex = 1.2) 
points(1:2, limits["LCL", ], pch = 118) 

abline(h = 0.23, lty = 3)
text(0.5, 0.24, "PRG = 0.23 mg/L", adj = 0, cex = 0.8)

mtext(paste("FOD =", stats.no.ND["N", ], "/", stats["N", ]), 
	at = 1:2, side = 1, line = 3)

legend(1.25, 0.07, c("outlier > 1.5 x IQR", "95% UCL", "Mean", "95% LCL"), 
	pch = c(1, 94, 8, 118), pt.cex = c(1, 1.2, 0.7, 1),
	bty = "n")

rm(TCE, Well, stats, stats.no.ND, box.list, limits)
```

# Example 9-3, pp. 9-9 to 9-12
```{r}
TCE  <- EPA.09.Table.9.1.TCE.df$TCE.mg.per.L
Well <- EPA.09.Table.9.1.TCE.df$Well

#####
# Frequency Histograms
#####

##
# Variation 1: Create histograms using the hist() function
par(mfrow = c(2, 1), mar = c(5, 4, 0, 1) + 0.1, oma = c(0, 0, 3, 0))
hist(TCE[Well == "Well.1"], col = "blue", ylim = c(0, 10),
	xlab = "Trichloroethene (mg/L) in Well 1", 
	ylab = "Frequency", main = "")
hist(TCE[Well == "Well.2"], col = "blue", ylim = c(0, 8),
	xlab = "Trichloroethene (mg/L) in Well 2", 
	ylab = "Frequency", main = "")
mtext("Figure 9-3. Frequency Histrograms of Trichloroethene by Well.", 
	outer=T, line = 1, cex = 1.25, font = 2)

# Variation 2: Force the x-axis to be the same for each histogram
xlim <- range(TCE, na.rm = T)
par(mfrow = c(2, 1), mar = c(5, 4, 0, 1) + 0.1, oma = c(0, 0, 3, 0))
hist(TCE[Well == "Well.1"], col = "blue", 
	xlim = xlim, ylim = c(0, 10),
	xlab = "Trichloroethene (mg/L) in Well 1", 
	ylab = "Frequency", main = "")
hist(TCE[Well == "Well.2"], col = "blue", 
	xlim = xlim, ylim = c(0, 8),
	xlab = "Trichloroethene (mg/L) in Well 2", 
	ylab = "Frequency", main = "")
mtext("Figure 9-3. Frequency Histrograms of Trichloroethene by Well.", 
	outer=T, line = 1, cex = 1.25, font = 2)

# Variation 3: Create histograms using the function 
# histogram in the lattice library.

library(lattice)
histogram(~ TCE.mg.per.L | Well, data = EPA.09.Table.9.1.TCE.df, 
	type = "count", layout = c(1, 2), 
	xlab = "Trichloroethene (mg/L)", ylab = "Frequency",
	main = "Figure 9-3. Frequency Histrograms of Trichloroethene by Well.")
	
#####
# Relative Frequency Histograms
#####

##
# Variation 1: Create relative frequency histograms using the hist() function
par(mfrow = c(2, 1), mar = c(5, 4, 0, 1) + 0.1, oma = c(0, 0, 3, 0))
hist(TCE[Well == "Well.1"], freq = F, 
	col = "blue", ylim = c(0, 15), 
	xlab = "Trichloroethene (mg/L) in Well 1", 
	ylab = "Relative Frequency (%)", main = "")
hist(TCE[Well == "Well.2"], freq = F, 
	col = "blue", ylim = c(0, 25),
	xlab = "Trichloroethene (mg/L) in Well 2", 
	ylab = "Relative Frequency (%)", main = "")
mtext("Figure 9-4. Relative Frequency Histrograms of Trichloroethene by Well.", 
	outer=T, line = 1, cex = 1.25, font = 2)

##
# Variation 2: Force the x-axis to be the same for each histogram
xlim <- range(TCE, na.rm = T)
par(mfrow = c(2, 1), mar = c(5, 4, 0, 1) + 0.1, oma = c(0, 0, 3, 0))
hist(TCE[Well == "Well.1"], freq = F, col = "blue", 
	xlim = xlim, ylim = c(0, 15),
	xlab = "Trichloroethene (mg/L) in Well 1", 
	ylab = "Relative Frequency (%)", main = "")
hist(TCE[Well == "Well.2"], freq = F, col = "blue", 
	xlim = xlim, ylim = c(0, 25),
	xlab = "Trichloroethene (mg/L) in Well 2", 
	ylab = "Relative Frequency (%)", main = "")
mtext("Figure 9-4. Relative Frequency Histrograms of Trichloroethene by Well.", 
	outer=T, line = 1, cex = 1.25, font = 2)

##
# Variation 3: Create relative frequency histograms using the function 
#              histogram in the lattice library.

library(lattice)
histogram(~ TCE.mg.per.L | Well, data = EPA.09.Table.9.1.TCE.df, 
	type = "percent", layout = c(1, 2), 
	xlab = "Trichloroethene (mg/L)", ylab = "Relative Frequency (%)",
	main = "Figure 9-4. Relative Frequency Histrograms of Trichloroethene by Well.")

rm(TCE, Well, xlim)
```

# Example 9-4, pp. 9-13 to 9-14
```{r}
As    <- EPA.09.Table.9.3.df$Arsenic.mg.per.L
As.DQ <- EPA.09.Table.9.3.df$Arsenic.Data.Qualifier

Hg    <- EPA.09.Table.9.3.df$Mercury.mg.per.L
Hg.DQ <- EPA.09.Table.9.3.df$Mercury.Data.Qualifier

ND.ind <- rep("", length(As))
ND.ind[As.DQ != "U" & Hg.DQ != "U"] <- "both detected"
ND.ind[As.DQ == "U" & Hg.DQ == "U"] <- "both non-detects"
ND.ind[As.DQ == "U" & Hg.DQ != "U"] <- "arsenic non-detect only"
ND.ind[As.DQ != "U" & Hg.DQ == "U"] <- "mercury non-detect only"
ND.ind <- factor(ND.ind, levels = c("both detected", 
	"both non-detects", "arsenic non-detect only", 
	"mercury non-detect only"))

# Summary statistics for Arsenic, setting non-detects to detection limit
summaryStats(As)

# Summary statistics for Mercury, setting non-detects to detection limit
summaryStats(Hg)


# NOTE: there is one (As,Hg) pair (0.01, 0.02) where both are detects
#       and      one (As,Hg) pair (0.01, 0.02) where both are non-detects
#       so you can't use a solid symbol for one or both or they will
#       not both show up!

lim <- range(As, Hg, na.rm = T)
pch.vec <- c(1, 4, 6, 2)
col.vec <- c("black", "red", "blue", "blue")
levels.ND.ind <- levels(ND.ind)
n.levels <- length(levels(ND.ind))

plot(As, Hg, type = "n", xlim = lim, ylim = lim, 
	xlab = "Arsenic (mg/L)", ylab = "Mercury (mg/L)",
	main = "Figure 9-5. Sctterplot of Arsenic with Mercury from Well 3")
for(i in 1:n.levels) {
	index <- ND.ind == levels.ND.ind[i]
	points(As[index], Hg[index], lwd = 2,
		pch = pch.vec[i], col = col.vec[i])
}
abline(lm(Hg ~ As))
mtext(paste("Pearson Correlation =", 
	round(cor(As, Hg, use="complete.obs"), 2)), line = 0.5)

legend(0.01, 0.3, levels.ND.ind, pch = pch.vec, 
	col = col.vec, lty = 0, lwd = 2, bty = "n")

rm(As, As.DQ, Hg, Hg.DQ, ND.ind, lim, 
	pch.vec, col.vec, levels.ND.ind, n.levels,
	i, index)
```

# Example 9-5, pp. 9-15 to 9-16
```{r}
As    <- EPA.09.Table.9.3.df$Arsenic.mg.per.L
As.DQ <- EPA.09.Table.9.3.df$Arsenic.Data.Qualifier

Hg    <- EPA.09.Table.9.3.df$Mercury.mg.per.L
Hg.DQ <- EPA.09.Table.9.3.df$Mercury.Data.Qualifier

Sr    <- EPA.09.Table.9.3.df$Strontium.mg.per.L
Sr.DQ <- EPA.09.Table.9.3.df$Strontium.Data.Qualifier


lim <- range(As, Hg, Sr, na.rm = T)
pch.vec <- c(1, 4, 6, 2)

cor.vec <- rep(as.numeric(NA), 3)

# First plot Hg vs. As

ND.ind <- rep("", length(As))
ND.ind[As.DQ != "U" & Hg.DQ != "U"] <- "both detected"
ND.ind[As.DQ == "U" & Hg.DQ == "U"] <- "both non-detects"
ND.ind[As.DQ == "U" & Hg.DQ != "U"] <- "As non-detect only"
ND.ind[As.DQ != "U" & Hg.DQ == "U"] <- "Hg non-detect only"
ND.ind <- factor(ND.ind, levels = c("both detected", 
	"both non-detects", "As non-detect only", 
	"Hg non-detect only"))

levels.ND.ind <- levels(ND.ind)
n.levels <- length(levels(ND.ind))

plot(As, Hg, type = "n", xlim = lim, ylim = lim, 
	xlab = "mg/L", ylab = "mg/L",
	main = "Figure 9-6. Coded Sctterplot of Well 3 Arsenic, Mercury, and Strontium")
for(i in 1:n.levels) {
	index <- ND.ind == levels.ND.ind[i]
	points(As[index], Hg[index], lwd = 2,
		pch = pch.vec[i], col = "blue")
}
abline(lm(Hg ~ As), lwd = 2, col = "blue")
cor.vec[1] <- round(cor(As, Hg, use="complete.obs"), 2)


# Now plot Sr vs. As

ND.ind <- rep("", length(As))
ND.ind[As.DQ != "U" & Sr.DQ != "U"] <- "both detected"
ND.ind[As.DQ == "U" & Sr.DQ == "U"] <- "both non-detects"
ND.ind[As.DQ == "U" & Sr.DQ != "U"] <- "As non-detect only"
ND.ind[As.DQ != "U" & Sr.DQ == "U"] <- "Sr non-detect only"
ND.ind <- factor(ND.ind, levels = c("both detected", 
	"both non-detects", "As non-detect only", 
	"Sr non-detect only"))

levels.ND.ind <- levels(ND.ind)
n.levels <- length(levels(ND.ind))

for(i in 1:n.levels) {
	index <- ND.ind == levels.ND.ind[i]
	points(As[index], Sr[index], lwd = 2,
		pch = pch.vec[i], col = "red")
}
abline(lm(Sr ~ As), lwd = 2, col = "red")
cor.vec[2] <- round(cor(As, Sr, use="complete.obs"), 2)


# Now plot Sr vs. Hg

ND.ind <- rep("", length(Hg))
ND.ind[Hg.DQ != "U" & Sr.DQ != "U"] <- "both detected"
ND.ind[Hg.DQ == "U" & Sr.DQ == "U"] <- "both non-detects"
ND.ind[Hg.DQ == "U" & Sr.DQ != "U"] <- "Hg non-detect only"
ND.ind[Hg.DQ != "U" & Sr.DQ == "U"] <- "Sr non-detect only"
ND.ind <- factor(ND.ind, levels = c("both detected", 
	"both non-detects", "Hg non-detect only", 
	"Sr non-detect only"))

levels.ND.ind <- levels(ND.ind)
n.levels <- length(levels(ND.ind))

for(i in 1:n.levels) {
	index <- ND.ind == levels.ND.ind[i]
	points(Hg[index], Sr[index], lwd = 2,
		pch = pch.vec[i], col = "green")
}
abline(lm(Sr ~ Hg), lwd = 2, col = "green")
cor.vec[3] <- round(cor(As, Sr, use="complete.obs"), 2)


legend(0.01, 0.3, c(paste(
	"Mercury (Y) vs. Arsenic (X), Cor =", 
		cor.vec[1]),
	paste("Strontium (Y) vs. Arsenic (X), Cor =", 
		cor.vec[2]), 
	paste("Strontium (Y) vs. Mercury (X), Cor =",
		cor.vec[3])
	),  
	col = c("blue", "red", "green"), pch = 1, 
	lty = 1, cex = 0.8, lwd = 2, bty = "n")


rm(As, As.DQ, Hg, Hg.DQ, Sr, Sr.DQ, 
	ND.ind, lim, 
	pch.vec, levels.ND.ind, n.levels,
	i, index, cor.vec)

# NOTE:  It is easier to see what is going on by using 
#        matrix scatterplots, rather than
#        a coded scatterplot.

# Matrix scatterplots with:
# non-detects plotted at detection limit, 
# duplitcate points overlaid, 
# no regression lines, and
# no special symbols for non-detects.

dum.df <- EPA.09.Table.9.3.df[, 
	c("Arsenic.mg.per.L", "Mercury.mg.per.L", "Strontium.mg.per.L")]
names(dum.df) <- c("As (mg/L)", "Hg (mg/L)", "Sr (mg/L)")

lim <- range(dum.df, na.rm = T)

pairs(dum.df, xlim = lim, ylim = lim)


# Matrix scatterplots with:
# non-detects plotted at detection limit, 
# duplitcate points jittered, 
# regression lines included, but
# no special symbols for non-detects,
# points plotted in upper panels,
# Pearson correlations shown in lower panels.

panel <- function(x, y, factor.x = 10, factor.y = 10, ...) {
	points.w.dups(x, y, 
		factor.x = factor.x, 
		factor.y = factor.y, ...)
	abline(lm(y~x))
}

panel.cor <- function(x, y, digits=2, prefix="Cor = ", 
	cex = 2 * par("cex"), ...)
{
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- cor(x, y, use = "pairwise.complete.obs")
    txt <- format(c(r, 0.123456789), digits=digits)[1]
    txt <- paste(prefix, txt, sep="")
    text(0.5, 0.5, txt, cex = cex)
}


set.seed(233)
pairs(dum.df, upper.panel = panel, 
	lower.panel = panel.cor, 
	xlim = lim, ylim = lim)


rm(dum.df, panel, panel.cor, lim)
```

# Example 9-6, pp. 9-17 to 9-20
```{r}
qqPlot(EPA.09.Table.9.4.nickel.vec, ylim = c(0, 1000),
	plot.pos.con = 0, add.line = T,
	ylab = "Nickel Concentration (ppb)",
	main = "Figure 9.7.  Normal Probability Plot")

qqPlot(EPA.09.Table.9.4.nickel.vec, dist = "lnorm", 
	plot.pos.con = 0, add.line = T,
	ylab = "log(nickel) Concentration log(ppb)",
	main = "Figure 9.8.  Probability Plot of Log Transformed Nickel Data")
```
